---
title: "ESM 244 lab 1"
output: html_document
---

```{r setup, include=TRUE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(here)
library(sf)
library(tmap)
```
cmd-shift-enter shortcut for running the current code chunk
cmd-enter shortcut for running the current line

### read in data

cmd-option-i shortcut for creating a code chunk
```{r}
sf_trees <- read_csv(here('data', 'sf_trees', 'sf_trees.csv'))
```

### Part 1: wrangling and ggplot review

```{r}
# method 1: group_by() %>% summarize()
sf_trees %>% 
  group_by(legal_status) %>% 
  summarize(tree_count = n())

# method 2
top_5_status <- sf_trees %>% 
  count(legal_status) %>% 
  drop_na(legal_status) %>% 
  rename(tree_count = n) %>% 
  relocate(tree_count) %>% 
  slice_max(tree_count, n = 5) %>% 
  arrange(-tree_count)
```

Make a graph of the top 5 from above:

```{r}
ggplot(data = top_5_status, aes(x = fct_reorder(legal_status, tree_count), y = tree_count)) +
  geom_col(fill = 'darkgreen') +
  labs(x = 'legal status', y = 'Tree count') +
  coord_flip() +
  theme_minimal()
```

**Example 2:**Only going to keep observations where legal status is 'permitted site' and caretaker is 'MTA, and store as `permitted_data_df`

shift-cmd-c to comment/uncomment quickly

```{r}
# sf_trees$legal_status %>% unique()
# unique(sf_trees$legal_status)

permitted_data_df <- sf_trees %>% 
  filter(legal_status == "Permitted Site", caretaker == "MTA")
```

**Example 3:** Only keep blackwood acacia trees, and then only keep certain columns

```{r}
blackwood_acacia_df <- sf_trees %>% 
  filter(str_detect(species, 'Blackwood Acacia')) %>% 
  select(legal_status, date, lat = latitude, lon = longitude)

### make a little graph of locations
ggplot(data = blackwood_acacia_df, aes(x = lon, y = lat)) +
  geom_point(color = 'darkgreen')
```
**Example 4:** use `tidyr::separate()`
```{r}
sf_trees_sep <- sf_trees %>% separate(species, into = c('spp_scientific', 'spp_common'), sep = ' :: ')
```

**Example 5:** use `tidyr::unite()`

```{r}
ex_5 <- sf_trees %>% 
  unite('id_status', tree_id, legal_status, sep = "_COOL_")
```

